\NAME: A framework for \\ Efficient Intrusion Detection\\ in the Internet of Things

### 2014-11-13 paper.tex
another round of improvements

When intrusion prevention fails, intrusion detection serves as a second layer
of defense. It looks for patterns \& anomalies that can indicate malicious
behavior. But supporting an adequate set of intrusion detection algorithms
imposes significant overhead on resource-restricted devices such as those
targeting the Internet of Things. Fusion of the algorithms, to optimize sharing
of resources, can alleviate this, yet, proves to be time-consuming, repetitive
and error-prone. To address this problem we introduce \NAME, a framework for
the development of efficient intrusion detection systems. It consists of a
domain specific language, to formally describe intrusion detection algorithms
on a functional level, and a code generator to produce more optimally organized
source code. A side-by-side comparison shows that generated code reduces memory
footprint, message passing overhead and execution time in comparison to a
straightforward sequential implementation of the same algorithms.

@@

The internet of things (IoT) holds a great potential to positively influence
our daily work and life through domotics, assisted living, e-health and
enhanced learning \cite{atzori2010internet}. But with this potential the IoT
also presents a significant threat: by opening our smart homes and personal
data to all these interconnected devices, we also open them to everyone who is
able to break the virtual locks (i.e.\ passwords, encryption keys,\dots) that
protect them. IoT devices typically have limited batteries, processing power
and memory. These properties make it hard to add any additional service,
including security.

@@

In classic networks, firewalls focus on the outer perimeter of the network,
filtering unwanted packets and protecting the entire internal network. But
attacks on flaws in services can pass unnoticed. This is where intrusion
detection (ID) steps in: an ID probe monitors all traffic that passes through
the firewall, looking for patterns and optionally, after detecting a malicious
pattern, alerting the firewall \cite{denning1987intrusion}.

@@

In wireless networks of resource constraint devices, which make up an
significant part of the IoT, it is not possible to have this single point in
the network that can oversee all traffic \cite{mishra2004intrusion}. Every
device has to implement its own lines of defense. But making the intrusion
detection system (IDS) a local service on each device requires local resources,
which are a scarce commodity for IoT devices.

Attackers have access to a large set of possible attack vectors
\cite{aschenbruck2012security}, ranging from the wireless network, over the
medium access control and routing, up to the device itself. Each level presents
different opportunities to manipulate data, eavesdrop, perform a form of denial
of service or access the network in a legitimate way. For each of these attacks
an algorithm needs to look for patterns or anomalies. An IDS for IoT devices
therefore requires a large number of implemented algorithms for an adequate
coverage.

@@

Current developments in ID focus on programming frameworks to structure the
implementation of algorithms \cite{valero2012di} and offer the required basic
functional components to implement them \cite{krontiris2008lidea}. The first
aspect that is missing, is that they don't offer a way to optimize the usage of
resources. A solution that supports the integration of multiple algorithms on
IoT devices needs to consider the impact on their resources and avoid
accumulating their impact.

@@

The first contribution of this paper is a framework that enables the formal
description of ID algorithms on a functional level. An automated code
generation process produces corresponding source code, for a given platform and
configuration. This way, the framework addresses the heterogeneous nature of
both the IoT and ID algorithms.

@@

The second contribution introduces functional code fusion (FCF) as a generation
paradigm to solve the problem of hard to reuse and combine ID algorithms. FCF
identifies common data and functions, and organizes code to eliminate redundant
iterations, tests and computations. Results show improved memory usage,
execution time and usage of the wireless radio, reducing energy consumption.

@@

The remainder of this paper proceeds as follows: section \ref{classification}
analyses and classifies ID algorithms. Section \ref{pattern} looks for patterns
in the implementation of these classes and details the ideas behind FCF\@.
Section \ref{design} describes the design we applied to construct the
framework, its domain specific language (DSL) and code generator. Section
\ref{evaluation} evaluates an implementation of the framework. Section
\ref{related} explores related work in the field of DSLs for ID and code
generation for the IoT\@. Finally, section \ref{conclusion} summarizes our
findings, draws conclusions and identifies topics for future work.

@@

Two aspects, currently missing from ID frameworks, that \NAME wants to cover
are optimization of resources and support for systematic reuse. This section
uses a classification to define the scope of \NAME, detailing which types of
algorithms are good candidates for the proposed approach.

Research into ID in wireless networks typically focuses on two major topics:
the devices as single entities and the network as a group of such devices. Both
are important: a group cannot make a decision without members that detect
malicious behavior and a group-based decision is often needed because nodes in
the network can miss certain events that could reveal intrusions. In a wireless
network, where not all participants are within each others' radio range, nodes
typically only have a partial view on the entire network. This prohibits them
to observe potentially crucial events in distant parts of the network. The fact
that resource constraint devices typically apply duty cycling to lower their
energy consumption, further increases this problem, introducing periods of time
where they even don't see local events.

@@

Because not all devices are always actively participating in the network, they
can miss clues that would otherwise lead them to detect intrusions. It is
hardly impossible for a single device to detect a complex attack by itself.
Therefore a second important research topic consists of cooperative algorithms
to combine information from single identities into a group-based decision about
alleged intrusions. We identify two ways to construct cooperative decision
making and focus on the way participants exchange information: broadcasting and
interactive.

@@

  \item[Broadcasting] can sometimes allow to implement a distributed,
  cooperative algorithm locally. The reputation-based detection algorithm
  introduced in \cite{ganeriwal2008reputation} illustrates this, based on the
  same principle of a watchdog also found in \cite{mishra2004intrusion}. Using
  this algorithm, nodes can exchange information about the reputation of other
  nodes, solely with broadcasts. Combining this information allows them to
  decide about their trust in a given node, based on more than their own, often
  partial, observations alone.

  \item[Interactive] communication allows algorithms to exchange information to
  reach a consensus. In \cite{krontiris2009cooperative}, the authors first
  present a theoretical foundation to analyze cooperative algorithms. Given
  these foundations they continue to present an algorithm consisting of 5
  phases. It essentially implements a voting system-based on the Guy Fawkes
  protocol \cite{anderson1998new} to identify an intruder. It enables the
  exchange of suspected intruder information and allows distributed
  authentication of those \emph{votes}. Based on these authenticated votes, the
  participants can make a distributed decision about the jointly identified
  intruder.

@@

Algorithms that fit this classification are typically platform independent.
They analyze network messages, looking for patterns that can reveal intrusions
and deal with the same information: other nodes in the network. To be able to
formally describe ID algorithms and target the heterogeneous IoT, \NAME
requires ID algorithms to expose these properties.

@@

\begin{figure}[ht]
\captionof{lstlisting}{ID Algorithm Pattern\label{alg:id-algo-pattern}}
\begin{algorithmic}[1]
  \Require{nodes, global storage for information about nodes}
  \Function{process\_message}{$msg$}
    \ForEach{$byte \in msg$} \label{alg:id-algo-pattern-loop1}
     \State \dots \Comment{analyze byte-sequences}
    \EndFor
    \Let{$nodes_x$}{value}  \Comment{optionally update node info}
    \State \Call{send}{$nodes_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
  \EndFunction
  \State
  \Function{do\_housekeeping}{}
    \ForEach{$node \in nodes$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
      \If{$node > \dots$} \Comment{validate recorded value}
        \State \dots \Comment{take actions}
        \State \Call{send}{$node$, ``info''} \Comment{communicate} \label{alg:id-algo-pattern-send2}
      \EndIf
    \EndFor
  \EndFunction
  \State
\end{algorithmic}
\end{figure}

We now extend the scope of the design pattern, to include its application in
the context of multiple algorithms. When implementing different algorithms, a
pattern such as described in listing \ref{alg:id-algo-application} will
typically appear.

@@

The pattern calls the different, linked implementations in a sequential way in
two situations: first, when the device observes new messages on the network
and, secondly, at a regular interval to allow for internal housekeeping, such
as threshold evaluation.

\subsection{Inherent Problems}
\label{pattern-problems}

These patterns can now help in identifing the problem areas that tend to result
in suboptimal code when combining multiple ID algorithms. Two issues arise from
the nested loops that originate from the combination of the ID algorithm
pattern and the application pattern: loops and common data. A third issue is
the use of the network.

@@

  \item[Common data] cross-cuts the different algorithms, that have to store
  information about nodes within its own scope. This results in a lot of
  duplicated information in memory.

@@

\bigskip
Two operational aspects are orthogonal to these problems and amplify them:

@@

\begin{figure}[ht]
  \centering
  \scalebox{.85}{
\begin{tikzpicture}[
  node distance=7mm,
  ,>=stealth,thick,black!50,text=black,
  every new ->/.style={shorten >=1pt}
]
  \node (dsl)  [resource]                                 {DSL};
  \node (cg)   [representation,right=of dsl,align=center] {Code\\Generator};
  \node (code) [resource,right=of cg]                     {Code};

  \draw [->] (dsl) -> (cg)   node [midway,above=25pt]              {formal};
  \draw [->] (cg)  -> (code) node [midway,above=15pt,align=center] {platform\\ + language};

@@

The DSL enables researchers to formally describe ID algorithms, without
focusing on any specific language or platform. A code generator can combine
such formal descriptions, avoiding the identified problems concerning loops,
common data and network access. It can also target different languages and
platforms, while reusing a software library that provides services including
parsing, network access wrapping,\dots

@@

\begin{figure}[ht]
  \centering
  \scalebox{.85}{
\tdplotsetmaincoords{110}{-10}
\begin{tikzpicture}[scale=1,tdplot_main_coords]
  % axis
  \draw[thick,->] (0,0,0) -- (7,0,0) node[anchor=north east,xshift=10mm]{$flexibility$};
  \draw[thick,->] (0,0,0) -- (0,7,0) node[anchor=north west]{$update\ cost$};
  \draw[thick,->] (0,0,0) -- (0,0,4) node[anchor=south]{$development$};
  % points
  \coordinate (O)        at (0,0,0);
  \coordinate (ideal)    at (6,0,0);
  \coordinate (vm)       at (2,2,2);
  \coordinate (modular)  at (4,4,4);
  \coordinate (reconfig) at (4,2,4);
  \coordinate (full)     at (6,6,4);
  \coordinate (cg)       at (6,4,2.5);
  % "lines"
  \draw[dashed] (2,0,0)    -- (2,2,0) -- (0,2,0);
  \draw[dashed] (4,0,0)    -- (4,4,0) -- (0,4,0);
  \draw[dashed] (6,0,0)    -- (6,6,0) -- (0,6,0);
  \draw[dashed] (vm)       -- (2,2,0);
  \draw[dashed] (modular)  -- (4,4,0);
  \draw[dashed] (full)     -- (6,6,0);
  \draw[dashed] (reconfig) -- (4,2,0);
  \draw[dashed] (cg)       -- (6,4,0);
  % "balls"
  \ball{6}{0}{0}{ideal}
  \ball{2}{2}{2}{virtual machine}
  \ball[left,xshift=-5pt]{4}{4}{4}{\shortstack{modular\\native image}}
  \ball{4}{2}{4}{\shortstack{reconfigurable\\virtual machine}}
  \ball[right,xshift=+5pt,yshift=+5pt]{6}{6}{4}{full native image}
  %\ball[right,xshift=+5pt,text=black]{6}{4.5}{3}{code generation};
  % 5. cg
  \tdplottransformmainscreen{6}{4}{2.5}
  \shadedraw[tdplot_screen_coords, ball color=gray,draw=black,shading angle=-90]
    (\tdplotresx,\tdplotresy) circle (0.20)
    node[right,xshift=+5pt,text=black,align=center,font=\footnotesize]{code generation\\+ functional code fusion};
  % "box around our solution" ;-)
  \tdplottransformmainscreen{0}{0}{0}
  \draw[tdplot_screen_coords] (4.95,1) rectangle (8.5,0.25);

@@

The \emph{ideal} situation is that were one ends up with a maximal flexibility
and no update nor development cost.

If we consider the trade off between flexibility and update cost, we can
distinguish two extremes: \emph{virtual machines} and \emph{full native
images}. Virtual machines allow for small updates due to their concise byte
code, but their instruction set and predefined capabilities restrict them at
the same time. A full native image on the other hand has unlimited flexibility,
but requires transfer of a larger image over the network.

Two intermediate alternatives are possible: \emph{modular native image} and
\emph{reconfigurable virtual machines}. Modular native images allow changes on
module level, limiting the update size, but the framework to support it is
static and limits the possibilities. Reconfigurable virtual machines take this
same approach, but apply it to virtual machines, allowing to \emph{upgrade} the
instruction set of the virtual machine.

\NAME starts from the full native image, allowing full flexibility, and aims to
reduce both the update and development cost. Firstly, by applying functional
code fusion, \NAME optimizes the code organization, which has a positive effect
on the code size and hence also on the update cost. Secondly, by enabling reuse
on a functional level and automating the writing phase, the development cost is
also lowered.

@@

The example of parsing network messages illustrates this. Each algorithm deals
with parsing in the same way: loop over the bytes in the payload and look for a
pattern that identifies what the algorithm is looking for. This is in fact a
technical implementation choice. The underlying functional goal is to react
when such a pattern is present in a message. The loop has \emph{activated} this
functionality, but the real goal is to \emph{passively} react on an event, not
\emph{actively} look for it. Event handling therefore can transform
constructions with loops to their functional meaning, allowing identification
and fusion of common functionality.

@@

All algorithms deal with information about neighboring devices. To centralize
this information \NAME provides a unified \ttt{nodes} concept as part of its
DSL and exposes it as an object to the implementer.

Algorithms can extend nodes with their own properties. This way only one
instance for every known device will be present in memory. Finally, nodes can
define a scope to execute a function in. This way, nodes also offer a way to
iterate all known devices, again without the need for a loop.

@@

Sending and receiving of network messages is tightly integrated with the
previously mentioned concepts of event handling and the nodes domain. An
underlying software framework will handle all interaction with the network,
triggering functions by means of events when parsing new messages and exposing
an API to send messages through the nodes domain. This approach gives the
framework complete control over communication, allowing optimized marshaling,
grouping,\dots

@@

A semantic model \cite{fowler2010domain}, tied closely to the DSL itself, can
fulfill this role. Semantic transformations evolve the model into a
syntax-oriented model, comparable to the idea of CodeDom. From there on,
syntactic transformations an emit source code. Figure \ref{fig:code-generation}
shows the process that \NAME implements.

@@

  \draw [->] (dsl) -> (sm)   node [midway,above=15pt]              {parse};
  \draw [->] (sm)  -> (cm)   node [midway,above=15pt,align=center] {semantic\\transform};
  \draw [->] (cm)  -> (code) node [midway,above=15pt]              {emit code};

  \draw [rounded corners,->]
        ($ (sm.east) - (5mm,5mm) $)
        -- ++(0,-.5)
        -| ($ (sm.west) + (5mm,-5mm) $)
           node [near start, below=1pt] {type inference};

@@

After parsing the DSL, it's imported into a semantic model. The minimal type
information present in the initial model allows inferring of the remaining
types. Through a series of semantic transformations, the model evolves from the
functional to a more technical syntactic model. This model is comparable to a
rich abstract syntax tree (AST) with an extensive set of language constructs.
Syntactic transformations now lower this level of expressiveness according to
the desired platform and language. In the end the syntactic model can be
directly emitted to the target language.

@@

The semantic model is the core of the solution. It offers a way to describe ID
algorithms in a way that allows semantic reinterpretation. This way,
reorganizing the functionality of different algorithms allows for more optimal
execution and memory usage. Hence the name \NAME of the framework: intrusion
detection functionality organization optimization.

@@

@@

Section \ref{pattern-problems} introduced the option to use a software library
to offload common implementation aspects, thus avoiding small discrepancies in
code and suboptimal use of shared resources. Section \ref{dsl-unified-msg}
illustrated this approach with a case for unified messaging: parsing of
incoming messages and grouping of outgoing messages.

Other common functionality also isn't fully generated. The execution of
functions based on a schedule or the implementation of higher level conceptual
functions, such as \ttt{now} and \ttt{sha1}, are calls into the standard
software library that supports the code generator.

Optimizing such algorithms is not within the scope of \NAME. It also gives
freedom of choice to the users of the framework to select appropriate
implementations for this standard functionality.

@@

\NAME reduces inefficiencies in execution, memory usage and network access, due
to the selection of separate ID algorithms. To achieve this, it avoids
constructions that are hard to identify and reorganize, such as loops and
common data. A DSL, focussing on event-handling, enforces these restrictions
and enables formally describing ID algorithms.

@@

This section discusses a prototype implementation and evaluates its
performance. We first present the architecture of this implementation and
consider the impact it has from a development point of view. Next we use this
implementation to quantify its merits using a side-by-side comparison of
generated code versus a manual implementation.

@@

@@

Figure \ref{fig:architecture} shows the architecture of the implementation. It
illustrates the generation of the DSL into compilable code, using the Nodes
domain information, a platform and language description and the native software
library. The applied grayscale indicates the components' platform dependency:
the DSL is platform and language independent and deals with subsystems, such as
the network, in an abstract way. The Nodes domain implements some of these
abstracted subsystems, but still in a platform and language independent way.
The platform and language descriptions define the target platform and language
and convert the abstract constructions in actually supported code. The software
library is standard native code, targeting the platform and language. Finally,
the code and image are artifacts of the automated generation and compilation
process. They aren't a product of manual work and therefore aren't directly
dependent on the platform.

@@

@@

To evaluate the prototype we chose to use a system based on the Atmel
ATMEGA1284p micro-controller \cite{datasheet:atmega1284p} and the Digi XBee S2
ZigBee module \cite{manual:xbee}. A basic library wrapped the technical calls
to the hardware components in more easy to use functions.

@@

We evaluated three criteria: the image size of the resulting compiled code, the
network usage in number of frames and bytes and the time required to perform
once cycle of the event loop. To do this, we collected these metrics for four
situations: without ID algorithms, with a watchdog, with reputation-tracking
and with both algorithms implemented.

The manual implementation implemented both algorithms as standalone modules and
sequentially called them from the base-application's event loop. Given the
\NAME descriptions of the algorithms, the code generator generated all four
cases.

@@

Thirdly, when turning to the goal to reduce the network usage, we see that both
the number of frames as well as the amount of bytes sent have dropped by 13 and
7\%.

Finally we see that both algorithms by itself add an equal amount of time to
the processing of a single event loop cycle, but when combined, the event loop
is about 7\% faster compared to the manual case.

@@

The STAT framework with its STATL
language \cite{eckmann2002statl,vigna2003designing} are not intrusion detection
specific, but through extensions they allow to model attack scenarios using
state machines. Our event-driven approach shares this same ideas.

>>>>

When intrusion prevention fails, intrusion detection serves as a second layer
of defense. It looks for patterns \& anomalies that can indicate malicious
behavior. But supporting an adequate set of intrusion detection algorithms
imposes significant overhead on resource-constraint devices such as those in
the Internet of Things. Fusion of the algorithms, to optimize sharing of
resources, can alleviate this, yet, proves to be time-consuming, repetitive and
error-prone. To address this problem \NAME provides a framework for the
development of efficient intrusion detection systems. It consists of a domain
specific language, to formally describe intrusion detection algorithms on a
functional level, and a code generator to produce more optimally organized
source code. A side-by-side comparison shows that generated code is smaller and
reduces message passing overhead, execution time and memory footprint in
comparison to a straightforward sequential implementation of the same
algorithms.

@@

The internet of things (IoT) holds a great potential to positively influence
our daily work and life through domotics, assisted living, e-health and
enhanced learning \cite{atzori2010internet}. But with this potential the IoT
also presents a significant threat: by opening our smart homes and personal
data to all these interconnected devices, we also open them to everyone who is
able to break the virtual locks that protect them (i.e.\ passwords, encryption
keys,\dots). IoT devices typically have limited batteries, processing power and
memory. These properties make it hard to add any additional service, including
security.

@@

In classic networks, firewalls focus on the outer perimeter of the network,
filtering unwanted packets and protecting the entire internal network. But
attacks on flaws in services can pass unnoticed. This is where intrusion
detection (ID) comes into play: an intrusion detection systems (IDS) monitors
all traffic that passes through the firewall, looking for patterns of malicious
activity, and optionally, after detecting such pattern, alerting the firewall,
allowing it to take corrective actions \cite{denning1987intrusion}.

@@

In wireless networks of resource-constraint devices, which make up a
significant part of the IoT, it is not possible to have this single point in
the network that can oversee all traffic \cite{mishra2004intrusion}. Every
device has to implement its own lines of defense. But making the IDS a local
service on each device requires local resources, which are a scarce commodity
for IoT devices.

Attackers have access to a large set of possible attack vectors
\cite{aschenbruck2012security}, ranging from the wireless network, over the
medium access control and routing, up to the device itself. Each level presents
different opportunities to manipulate data, eavesdrop, perform a form of denial
of service or even access the network in a legitimate way. For each of these
attacks an algorithm needs to look for patterns or anomalies. An IDS for IoT
devices therefore requires a large number of implemented algorithms for an
adequate coverage.

@@

Current developments in ID focus on programming frameworks to structure the
implementation of algorithms \cite{valero2012di} and offer the required basic
functional components to implement them \cite{krontiris2008lidea}. The first
aspect that is missing, is that they don't offer a way to optimize the usage of
resources. A solution that supports the integration of multiple algorithms on
IoT devices needs to avoid accumulating their impact on the available resources.

@@

The first contribution of this paper is a framework called \NAME. It enables
the formal description of ID algorithms on a functional level and provides a
code generator to produces corresponding source code for a given platform and
configuration. This way, \NAME addresses the heterogeneous nature of both the
IoT and ID algorithms.

@@

The second contribution introduces functional code fusion (FCF) as a source
code generation paradigm to solve the problem of hard to reuse and combine ID
algorithms. FCF identifies common data and functions, and organizes code to
eliminate redundant iterations, tests and computations. Results show improved
image size, usage of the wireless radio, execution time and memory usage.

@@

The remainder of this paper proceeds as follows: section \ref{classification}
analyses and classifies ID algorithms. Section \ref{pattern} looks for patterns
in the implementation of these classes. Section \ref{problems} identifies the
problems arising from these patterns. Section \ref{design} describes the design
we applied to construct \NAME, its domain specific language (DSL) and code
generator. Section \ref{evaluation} discusses and evaluates an implementation.
Section \ref{related} explores related work in the field of DSLs and code
generation. Finally, section \ref{conclusion} summarizes our findings, draws
conclusions and identifies topics for future work.

@@

Two aspects, currently missing from ID frameworks, that \NAME wants to address,
are optimization of resources and support for systematic reuse. This section
uses a classification to define the scope of \NAME, detailing which types of
algorithms are good candidates for the proposed approach.

Research into ID in wireless networks typically focuses on two major topics:
the devices as single entities and the network as a group of such devices. Both
are important: a group cannot make a decision without members that detect
malicious behavior and a group-based decision is often needed because devices
in the network can miss certain events that could reveal intrusions. In a
wireless network, where not all participants are within each others' radio
range, wireless devices typically only have a partial view on the entire
network. This prohibits them to observe potentially crucial events in distant
parts of the network. The fact that resource-constraint devices typically apply
duty cycling to lower their energy consumption, further increases this problem,
introducing periods of time where they even don't observe local events.

@@

Because not all devices are always actively participating in the network, they
can miss events that would otherwise lead them to detect intrusions. It is
hardly impossible for a single device to detect a complex attack by itself.
Therefore a second important research topic consists of cooperative algorithms
to combine information from single identities into a group-based decision about
alleged intrusions. We identify two ways to construct cooperative decision
making and focus on the way participants exchange information: broadcast and
interactive communication.

@@

  \item[Broadcast communication] can allow implementing a distributed,
  cooperative algorithm locally. The reputation-based detection algorithm
  introduced in \cite{ganeriwal2008reputation} illustrates this. It uses the
  same principle of a watchdog also found in \cite{mishra2004intrusion}. Using
  this algorithm, devices can exchange information about the reputation of
  other devices, solely using one-way broadcast messages. Combining this
  information allows them to decide about their trust in a given device, based
  on more than their own, often partial, observations.

  \item[Interactive communication] allows algorithms to exchange information to
  reach a consensus. In \cite{krontiris2009cooperative}, the authors first
  present a theoretical foundation to analyze cooperative algorithms. Given
  these foundations they continue to present an algorithm consisting of five
  phases, which essentially implement a voting system based on the Guy Fawkes
  protocol \cite{anderson1998new}, to identify an intruder. It enables the
  exchange of suspected intruder information and allows for distributed
  authentication of those \emph{votes}. Based on these authenticated votes, the
  participants can make a distributed decision about the jointly identified
  intruder.

@@

Algorithms that fit this classification are typically platform independent.
They analyze network messages, looking for patterns that can reveal intrusions
and all deal with the same information: other devices in the network. To be
able to formally describe ID algorithms and target the heterogeneous IoT, \NAME
requires ID algorithms to expose these properties.

@@

\begin{figure}[ht]
\captionof{lstlisting}{ID Algorithm Pattern\label{alg:id-algo-pattern}}
\begin{algorithmic}[1]
  \Require{devices: global storage for device information}
  \Function{process\_message}{$msg$}
    \ForEach{$byte \in msg$} \label{alg:id-algo-pattern-loop1}
     \State \dots \Comment{analyze byte-sequences}
    \EndFor
    \Let{$devices_x$}{value}  \Comment{optionally update device info}
    \State \Call{send}{$device_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
  \EndFunction
  \State
  \Function{do\_housekeeping}{}
    \ForEach{$device \in devices$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
      \If{$device > \dots$} \Comment{validate recorded value}
        \State \dots \Comment{take actions}
        \State \Call{send}{$device$, ``info''} \Comment{communicate} \label{alg:id-algo-pattern-send2}
      \EndIf
    \EndFor
  \EndFunction
  \State
\end{algorithmic}
\end{figure}

Applying the pattern in the context of multiple algorithms, extends its scope
and a pattern such as described in listing \ref{alg:id-algo-application} will
typically appear.

@@

The pattern calls the different, linked implementations in sequence in two
situations: first, when the device observes new messages on the network and,
secondly, at a regular interval to allow for internal housekeeping, such as
threshold evaluation.

\section{Inherent Problems}
\label{problems}

These patterns can now help in identifying the problem areas that tend to
result in suboptimal code when combining multiple ID algorithms. Besides these
code-level problems, operational problems also hinder the implementation of an
IDS for IoT devices.

Two issues arise from the nesting of loops that originate from the combination
of the ID algorithm pattern and the application pattern: loops and common data.
A third issue is the use of the network.

@@

  \item[Common data] cross-cuts the different algorithms, that have to store
  information about devices within its own scope. This results in a lot of
  duplicated information in memory.

@@

Two operational aspects are orthogonal to these problems and amplify them:

@@

\begin{figure}[ht]
  \centering
  \scalebox{.85}{
\begin{tikzpicture}[
  node distance=7mm,
  ,>=stealth,thick,black!50,text=black,
  every new ->/.style={shorten >=1pt}
]
  \node (dsl)  [resource]                                          {DSL};
  \node (cg)   [representation,right=of dsl,align=center]          {Code\\Generator};
  \node (code) [resource,right=of cg]                              {Code};
  \node (lib)  [resource,below=of code,align=center,yshift=0.5cm]  {Software\\Library};

  \draw [->] (dsl) -> (cg)   node [midway,above=25pt]              {intent};
  \draw [->] (cg)  -> (code) node [midway,above=15pt,align=center] {platform\\ + language};

  \draw [rounded corners,->]
        ($ (lib.west) $)
        -- ++(-0.5cm, 0cm)
        -- ++( 0cm,   1cm)
        -- ($ (code.west) + (0cm,-0.22cm) $);

@@

The DSL enables researchers to formally describe ID algorithms, without
focusing on any specific language or platform. A code generator can avoid the
identified problems concerning loops, common data and network access, as the
formal description provides the underlying intent, not the technical
implementation. It can also target different languages and platforms, while
reusing a software library that provides services including parsing, network
access wrapping,\dots

@@

\begin{figure}[hb]
  \centering
  \scalebox{.85}{
\tdplotsetmaincoords{110}{-10}
\begin{tikzpicture}[scale=1,tdplot_main_coords]
  % axis
  \draw[thick,->] (0,0,0) -- (7,0,0) node[anchor=north east,xshift=10mm]{$flexibility$};
  \draw[thick,->] (0,0,0) -- (0,7,0) node[anchor=north west]{$update\ cost$};
  \draw[thick,->] (0,0,0) -- (0,0,4) node[anchor=south]{$development$};
  % points
  \coordinate (O)        at (0,0,0);
  \coordinate (ideal)    at (6,0,0);
  \coordinate (vm)       at (2,2,2);
  \coordinate (modular)  at (4,4,4);
  \coordinate (reconfig) at (4,2,4);
  \coordinate (full)     at (6,6,4);
  \coordinate (cg)       at (6,4,2.5);
  % "lines"
  \draw[dashed] (2,0,0)    -- (2,2,0) -- (0,2,0);
  \draw[dashed] (4,0,0)    -- (4,4,0) -- (0,4,0);
  \draw[dashed] (6,0,0)    -- (6,6,0) -- (0,6,0);
  \draw[dashed] (vm)       -- (2,2,0);
  \draw[dashed] (modular)  -- (4,4,0);
  \draw[dashed] (full)     -- (6,6,0);
  \draw[dashed] (reconfig) -- (4,2,0);
  \draw[dashed] (cg)       -- (6,4,0);
  % "balls"
  \ball{6}{0}{0}{ideal}
  \ball{2}{2}{2}{virtual machine}
  \ball[left,xshift=-5pt]{4}{4}{4}{\shortstack{modular\\native image}}
  \ball{4}{2}{4}{\shortstack{reconfigurable\\virtual machine}}
  \ball[right,xshift=+5pt,yshift=+5pt]{6}{6}{4}{full native image}
  %\ball[right,xshift=+5pt,text=black]{6}{4.5}{3}{code generation};
  % 5. cg
  \tdplottransformmainscreen{6}{4}{2.5}
  \shadedraw[tdplot_screen_coords, ball color=gray,draw=black,shading angle=-90]
    (\tdplotresx,\tdplotresy) circle (0.20)
    node[right,xshift=+5pt,text=black,align=center,font=\footnotesize]{code generation\\+ functional code fusion};
  % "box around our solution" ;-)
  \tdplottransformmainscreen{0}{0}{0}
  \draw[tdplot_screen_coords] (4.95,1) rectangle (8.5,0.25);

@@

The \emph{ideal} situation is that were one ends up with a maximal flexibility
and no update nor development cost. \NAME starts from the full native image,
allowing full flexibility, and aims to reduce both the update and development
cost: first, by applying functional code fusion, \NAME optimizes the code
organization, which has a positive effect on the code size and hence also on
the update cost. Second, by enabling reuse on a functional level and automating
the writing phase, the development cost is also lowered.

@@

The example of parsing network messages illustrates this. Each algorithm deals
with parsing in the same way: loop over the bytes in the payload and look for a
pattern that identifies what the algorithm is looking for. This is in fact a
technical implementation choice. The underlying functional intent is to react
when such a pattern is present in a message. The loop has \emph{activated} this
functionality, but the real goal is to \emph{passively} react on an event, not
\emph{actively} look for it. Event handling therefore can transform
constructions with loops to their functional meaning, allowing identification
and fusion of common functionality.

@@

All algorithms deal with information about other devices in the network. To
centralize this information \NAME provides a unified \ttt{nodes} concept as
part of its DSL and exposes it as an object to the implementer.

Algorithms can extend nodes with their own properties. This way, only one
instance for every known device will be present in memory. Finally, nodes can
define a scope for a function to execute in. This way, nodes also offer a way
to i.e.\ iterate all known devices, again without the need for a loop.

@@

Sending and receiving of network messages is tightly integrated with the
previously mentioned concepts of event handling and the Nodes domain. An
underlying software library can handle all interaction with the network,
triggering functions by means of events when parsing new messages and exposing
an API to send messages through the Nodes domain. This approach gives the
framework complete control over communication, allowing optimized marshaling,
grouping,\dots

@@

A semantic model \cite{fowler2010domain}, tied closely to the DSL itself, can
fulfill this role. Semantic transformations evolve the model into a more
syntax-oriented model, comparable to the idea of CodeDom. From there on,
syntactic transformations can further evolve the model into source code. Figure
\ref{fig:code-generation} shows the process that \NAME implements.

@@

  \draw [->] (dsl) -> (sm)   node [midway,above=15pt]              {parse};
  \draw [->] (sm)  -> (cm)   node [midway,below=10pt,align=center] {semantic\\transform}
                             node [midway,above=15pt,align=center] {FCF};
  \draw [->] (cm)  -> (code) node [midway,above=15pt]              {emit code};

  \draw [rounded corners,->]
        ($ (sm.east) - (5mm,5mm) $)
        -- ++(0,-.5)
        -| ($ (sm.west) + (5mm,-5mm) $)
           node [near start, below=5pt] {type inference};

@@

@@

The semantic model is the core of the solution. It offers a way to describe ID
algorithms in a way that allows semantic reinterpretation through functional
code fusion. This way, reorganizing the functionality of different algorithms
allows for more optimal execution and memory usage. Hence the name \NAME of the
framework: intrusion detection functionality organization optimization.

@@

One example of a semantic transformation is type inference, which supports the
platform independence of the framework.

@@

A software library allows offloading of common implementation aspects, thus
avoiding small discrepancies in code and suboptimal use of shared resources.
Section \ref{dsl-unified-msg} illustrated this approach with a case for unified
messaging: parsing of incoming messages and grouping of outgoing messages.

Other common functionality also isn't generated as-is. The execution of
functions based on a schedule or the implementation of higher level conceptual
functions, such as \ttt{now} and \ttt{sha1}, are calls into the standard
software library that supports the code generator.

Optimization of such algorithms is not within the scope of \NAME. It also gives
freedom of choice to the users of the framework to select appropriate
implementations for this standard functionality.

@@

\NAME reduces inefficiencies in network access, execution and memory usage,
that occur due to a selection of ID algorithms. To achieve this, it avoids
constructions that are hard to identify and reorganize, such as loops and
common data. A DSL, focussing on event-handling, enforces these restrictions
and enables formally describing ID algorithms.

@@

@@

This section discusses a prototype implementation and evaluates its
performance. We first present the architecture of this implementation and
consider the impact it has from a development point of view. Next we use this
implementation to quantify its merits using a side-by-side comparison of
generated code versus a manual implementation. We consider experimental results
regarding the image size, usage of the network and execution time of the
event-loop. We also take a more analytical approach and look at the handling of
common data and the size of the source code.

@@

Figure \ref{fig:architecture} shows the architecture of the implementation.

@@

It illustrates the generation of the DSL into compilable code, using the Nodes
domain information, a platform and language description and the native software
library. The applied grayscale indicates the components' platform dependency:
the DSL is platform and language independent and deals with subsystems, such as
the network, in an abstract way. The Nodes domain implements some of these
abstracted subsystems, but still in a platform and language independent way.
The platform and language descriptions define the target platform and language
and convert the abstract constructions in actually supported code. The software
library is standard native code, targeting the platform and language. Finally,
the code and image are artifacts of the automated generation and compilation
process. They aren't a product of manual work and therefore aren't directly
dependent on the platform.

@@

To evaluate the prototype we chose to use a system based on the Atmel
ATMEGA1284p micro-controller and the Digi XBee S2 ZigBee module. A basic
library wrapped the technical calls to the hardware components in more easy to
use functions.

@@

We evaluated three criteria: the image size of the resulting compiled code, the
network usage in number of packets and bytes and the time required to perform
once cycle of the event loop. To do this, we collected these metrics for four
situations: without ID algorithms, with a watchdog, with reputation-tracking
and with both algorithms implemented.

The manual implementation realized both algorithms as standalone modules and
sequentially called them from the base-application's event loop. Given the
\NAME descriptions of the algorithms, the code generator generated all four
cases.

@@

Thirdly, when turning to the goal to reduce the network usage, we see that both
the number of packets as well as the amount of bytes sent have dropped by 13
and 7\%.

Finally, focusing on the execution time, both algorithms by themselves add an
equal amount of time to the processing of a single event loop cycle. But when
combined, the event loop is about 7\% faster compared to the manual case.

@@

The STAT framework with its STATL language
\cite{eckmann2002statl,vigna2003designing} are not intrusion detection
specific, but through extensions they allow to model attack scenarios using
state machines. Our event-driven approach shares these same ideas.
